Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-05-11 paper.tex
correction of main part finished

In supernovae remnants and on scales of galaxies, there are convincing
evidences of non-thermal populations. Especially, in the MW, the
cosmic rays are observed directly as well as indirectly through radio,
X-ray, and gamma-ray emission. On larger scales of the order of few
100~kpc up to Mpc, there are currently a vast number of observations
of radio emission coming from radio mini halos in the center of
cooling flow clusters, radio relics in the periphery of clusters
\cite{2004rcfg.proc..335K} as well as smooth giant radio halos that
have been observed in more than 50 clusters
\cite{2003ASPC..301..143F,2008SSRv..134...93F}. This type of emission
is expected in clusters since the formation process of a galaxy
cluster is a very energetic processes that induces both turbulence as
well as frequently occurring merging and accretion shocks which both
are thought to accelerate relativistic non-thermal protons and radio
emitting electrons to high energies. Thus most of the observed radio
emission in clusters is expected to emerge from cosmic-ray electron
induced radio synchrotron emission. The precise origin of these
electrons, in especially relics and giant radio halos, is still not
settled. One possible scenario for the production of these electrons
are hadronic CR interactions with ambient gas protons which results in
the production of both charged and neutral pions that decay into
electrons, neutrinos, and gamma-rays (see
\cite{1980ApJ...239L..93D,1982AJ.....87.1266V,1999APh....12..169B,2000A&A...362..151D,2001ApJ...559...59M,2003MNRAS.342.1009M,2004A&A...413...17P,2004MNRAS.352...76P,2008MNRAS.385.1211P,2008MNRAS.385.1242P,2009JCAP...09..024K,2010MNRAS.401...47D,2010MNRAS.407.1565D,2010ApJ...722..737K,2010arXiv1011.0729K,2011A&A...527A..99E}
). \footnote{An alternative scenario is the second order turbulent
  re-acceleration through the interaction of a previously injected
  relativistic population of electrons by supernova driven winds or
  active galactic nuclei with plasma waves and
  magneto-turbulence.\cite{1987A&A...182...21S,1993ApJ...406..399G,2004MNRAS.350.1174B,2005MNRAS.363.1173B,2007MNRAS.378..245B,2011MNRAS.410..127B,2009A&A...507..661B}}
Supporting evidence comes from the smoothness of the extended radio
emission that often resembles that observed in thermal
X-rays. Especially since the long cooling time of cosmic ray protons
(CRps) allows for an extended and large population of CRps to build up
over the cluster history \citep{1996SSRv...75..279V,
  1997ApJ...477..560E, 1997ApJ...487..529B}. The production of these
secondaries depend both on the gas and CRp densities in the cluster,
where the CRp density roughly traces the gas outside the core regime
and is slightly enhanced in the center.  This density scaling implies
that clusters are great targets for Cherenkov telescopes with a high
sensitivity for the central parts of nearby clusters. Detecting the
cluster gamma-ray emission is crucial in this respect as it
potentially provides the unique and unambiguous evidence of CRp
populations in clusters through observing the $\pi^0$ bump at about
100~MeV in the spectra.

We adopt the universal spectral and spatial gamma-ray model developed
by Pinzke \& Pfrommer \cite{2010MNRAS.409..449P} to estimate the
emission from decaying $\pi^0$:s that dominates over the IC emission
from primary and secondary electrons above 100~MeV in clusters. The
gamma-ray formalism was derived from high resolution simulations of
clusters of galaxies that included radiative hydrodynamics, star
formation, supernova feedback, and followed CRp physics using a novel
formulation that trace the most important injection and loss processes
self-consistently while accounting for the CRp pressure in the
equation of motion
\cite{2008A&A...481...33J,2007A&A...473...41E,2006MNRAS.367..113P}.
We note that the overall normalization of the CRp and gamma-ray
distribution scales with the maximum acceleration efficiency at
structure formation shock waves. Following recent observations at
supernova remnants \cite{2009Sci...325..719H} as well as theoretical
studies \cite{2005ApJ...620...44K}, we adopt a realistic value of this
parameter and assume that 50\% of the dissipated energy at strong
shocks is injected into CRps while this efficiency rapidly decreases
for weaker shocks \cite{2007A&A...473...41E}. As a caveat we note that
the analytic model is based on simulations that only considered
advective transport of CRs by turbulent gas motions which produces
centrally enhanced profiles. However active CR transport such as CR
diffusion and streaming tends to drive the CR radial profiles towards
being flat, with equal CR number density everywhere. While the CR
streaming is then usually larger than typical advection velocities and
becomes comparable or lower than this only for periods with trans- and
supersonic cluster turbulence during a cluster merger. As a
consequence a bimodality of the CR spatial distribution results with
merging (relaxed) clusters showing a centrally concentrated (flat) CR
energy density profile \cite{2011A&A...527A..99E}.  This translates
into a bimodality of the expected diffuse radio and gamma-ray emission
of clusters, since more centrally concentrated CR will find higher
target densities for hadronic CR proton interactions
\cite{2011A&A...527A..99E}. As a result of this, relaxed clusters
could have a reduced gamma-ray luminosity by up to a factor of five.

@@

We focus in this section on the spectral distribution of gamma-rays
from clusters.  In the leptophilic model, DM annihilation radiation
includes final state radiation and give rise to substantial amounts of
electrons and positrons that IC upscatter background radiation fields
to high energies. We also consider a few supersymmetric DM models with
a high gamma-ray yield in the form of continuum emission and IC
induced emission. In addition to the annihilating DM, we estimate the
gamma-ray flux induced by shock accelerated CRps.

In Fig.~\ref{fig:diff_BM} we show the differential flux from the
Fornax cluster which is one of the best clusters for indirect DM
searches due to its relative high DM induced fluxes and low CR induced
fluxes. We derive the emission from four different supersymmetric BM
models and the emission induced by CRps. Comparing this emission to
the differential gamma-ray upper limits set by Fermi-LAT after 18
months of observations \cite{2010ApJ...717L..71A}, we find that the
upper limits are not violated. The predicted DM flux that is dominated
by the continuum emission from $\Kp$ and $\Ip$ models (shown in the
left panels) are close to the upper limits and can potentially be
probed by the Fermi 3 year data. Furthermore, the gamma-ray signal
induced by CRps is expected to be about a factor 10 below the DM
continuum flux from the $\Kp$ and $\Ip$ BM models around
10~GeV. However, the IC emission from upscattered CMB and SD photons
is at least a factor 10 lower than the expected flux from CRps above
100~MeV, making it very hard to distinguish such a signal from the
foreground due to the similar spectral index. For energies below
100~MeV we expect IC emission by primary shock accelerated electrons
to be dominating \cite{2010MNRAS.409..449P} over the supersymmetric DM
induced leptons. Hence the IC emission from supersymmetric DM in
clusters can for DM benchmark models be safely neglected compared to
the CRps and DM continuum emission.

@@

{\bf Anders: I got that far with the results, please polish the reminder up to population studies!}

\del{CONSIDER REWRITING, START WITH PHYSICS, THEN DISCUSS DOMINATING
CONTRIBUTIONS, PEAKS, AND SPECTRAL INDICES}

One of the main contributions to the expected extended gamma-ray flux
is coming from the IC upscattered photons on various photon background
fields. Depending on the spatial and spectral distribution of
electrons and positrons as well as the photon background field, the
resulting spectral distributions of gamma-ray can differ
greatly. Hence it is interesting to understand which spectral features
and energy regimes that are dominating for the different DM models.

If the enhancement due to substructures in clusters is significant,
then the distribution of electrons and positrons follows the radial
profile of substructures outside the cluster center where the smooth
DM density profile is dominating. In addition, the cooling of the
steady state electron and positron distribution is dominated by SD
photons in the center. Especially since the SD distribution is
centrally peaked and the substructure distribution peaks in the
outskirts around $\rvir$, the overlap between electrons/positrons and
SD photons is small. This results in a suppression of the IC
upscattered SD photons relative to the IC upscattered CMB photons, the
FSR, and the continuum emission. However, if substructures are only
marginally dominating over the smooth distribution in a cluster, the
SD component becomes more important, especially the flux from IC
upscattered CMB photons is small for LP models in the 100~GeV to 1~TeV
energy regime and in the 1~GeV to 100~GeV energy regime for BM models.

In Fig.~\ref{fig:IR_comp} we show the total IC emission as well as its
individual contributions from different IC upscattered radiation
fields. The left panel show the gamma-ray emission from the LP model
and the right panel from the $\Kp$ BM model, and for comparison we
overplot the emission expected from the CRps. Due to the flat electron
and positron spectrum resulting from the LP DM model and the smaller
mean energy of the CMB compared to the SD, we find that the
upscattered CMB photons dominate the total DM IC emission in the
energy regime below 100~GeV, while the SD dominate above this
energy. For the BM models this transition is shifted towards a smaller
energy since the electron and positron spectrum has a steep spectral
distribution that suppresses low energy photon fields. \del{This
  suppression is also seen in the relative normalization of the flux
  where the IC from both CMB and dust photons are lower compared to
  the starlight simply because the starlight has a higher energy,
  hence it is upscattered by more abundant low energy electrons and
  positrons (c.f. Fig.~\ref{fig:q_DM}).} At the highest gamma-ray
energies of about 100~GeV and above, the IC from starlight steepens
because it both probes the high energy tail of the electrons as well
as suffer from the KN suppression. It should be noted that if we
remove the boost from substructures, the density profile of electrons
and positrons is more centrally peaked, and the relative importance
of the IC upscattered SD increases by a factor $\sim 30$. In
addition, the IC upscattered CMB light would suffer more from SD
cooling in the center as most of its flux resides from the outer parts
when substructures are included. For comparison, we include the
contribution of IC upscattered SD photons and to bracket the
uncertainty in the SD model, with a small amount of dust. In this low
dust model we have reduced the energy in dust by a factor 10. However,
we find that the IC SD flux only decreases by a factor 3 since the
cooling of the electrons and positrons also decreases. Note that for a
smaller boost from substructures, the effect from cooling is smaller.
\del{Even though the SD component seems to be suppressed compared to
  the CMB at energies around 100~MeV, it has a relative large impact
  on the steady state electron spectrum. If we remove the SD cooling,
  we find a factor two larger IC flux at low energies for a cluster of
  the size of Fornax. For more massive clusters we expect a larger IC
  flux from the SD component, but a smaller contribution from the CMB
  because of the more effective cooling.}

Because of the large boost factor for the LP model ($\sim 10^4$), we
overproduce the upper gamma-ray limit in the $1-10$~GeV energy
interval set by Fermi-LAT by about a factor 100. This strongly
constrain the boost from both substructures and the Sommerfeld
enhancement. Additional, these constraints might improve with future
more sensitive Cherenkov telescopes such as CTA. However, considering
that an indirect detection of DM in clusters relies on the boost from
substructures whose main contribution comes from the periphery, we
conclude that the wide angular extent of clusters on the sky in
gamma-rays suggest that these sources are not ideal for Cherenkov
telescopes due to the loss of sensitivity outside the small PSF
{\bf BETTER CHECK THIS WITH SOME EXPERT}.

@@

It is also interesting to compare the total contribution from the LP
model, the brightest BM model ($\Kp$), and the CRp induced
emission. In Fig.~\ref{fig:flux_int} we show the integrated flux from
Fornax for our different gamma-ray models and compare it to the
unresolved integrated flux upper limit on Fornax set by Fermi-LAT
where they averaged the flux over the energy range $0.2-100\,\gev$
assuming a spectral index of 2. Again we are overproducing the upper
limits, although only with a factor 10 which is about a order
magnitude less constraining than the differential flux in the energy
range $1-10$~GeV. From the figure one clearly see that the LP model is
dominating the entire gamma-ray energy range up to the DM mass of this
model of about 1~TeV. Finally, we note that the flux the DM BM $\Kp$
model is above the predicted emission from the CRps in the $1-100$~GeV
energy regime accessible both to Fermi and Cherenkov
telescopes. Hence, in the absence of LP DM the prospects for detecting
the $\Kp$ BM model over the expected gamma-ray background induced by
CRps look promising.

@@

We continue by comparing the estimated differential flux from Fornax
to three other clusters in Fig.~\ref{fig:clu_comp}; the close by and
well studied Virgo cluster, the X-ray bright Perseus clusters, and the
massive merging Coma cluster. We find high DM induced gamma-ray fluxes
from both the Fornax and Virgo clusters, which highlight them as
promising targets for indirect DM searches. Note, however, that we
have used spatially unresolved upper limits, which is quite optimistic
for the angular extended Virgo cluster. It is quite striking how
constraining the upper limits for the Fornax cluster are compared to
the other clusters, especially in the $1-10$~GeV energy regime where
Fermi-LAT is most sensitive due to the maximized effective area. The
upper limits for Virgo and Perseus are unfortunately background
contaminated by AGN activity from M87 and NGC 1267, respectively, and
do not gain much from the increased sensitivity. One should also note
that the high CRp induced flux from both Coma and Perseus is of the
same magnitude as the DM flux from LP models, making it difficult for
indirect DM searches in those clusters. Conversely is the Fornax
cluster a great target for indirect DM studies because the relative
low gamma-ray flux from CRps, absents of an active AGN, and high DM
gamma-ray flux. Actually, the Fermi-LAT 3 year data will be able to
constrain the smallest mass of substructures using the DM BM
models. We have also learned that the closest clusters are superior
for DM searches compared to further away clusters, even though the
close by clusters have a mass that is much smaller. In fact, we can
understand this by looking at the luminosity-to-mass scaling
relations. The gamma-ray luminosity from the smooth density profile is
given by \cite{2009PhRvL.103r1302P}
\begin{equation}
L_{\gamma,\sm} \propto \int \dd V \rho(r)^2 \propto \frac{c^3}
{\left[\log\left(1+c\right)-c/(1+c)\right]} \propto \mvir^{0.83}\,,
\end{equation}
and the total DM luminosity including boost factors for the LP and
the BM models are given by
\begin{eqnarray}
\label{eq:LP_scaling}
L_{\gamma,\rmn{LP}} &=& L_{\gamma,\rmn{smLP}} \B_\rmn{LP} \propto \mvir^{0.73}\qquad
\,\,\,\,\,\,\rmn{and}\\
\label{eq:BM_scaling}
L_{\gamma,\rmn{BM}} &=& L_{\gamma,\rmn{smBM}} \B_\rmn{BM} \propto  \mvir^{1.06}\,,\qquad
\rmn{where} \\
\B_\rmn{LP}&=&\B_\sfe \B_\sub \propto \mvir^{-1/3}\mvir^{0.23} \propto
 \mvir^{-0.1}\,,\\
\B_\rmn{BM}&=&\B_\sub \propto \mvir^{0.23}\,.
\end{eqnarray}
Hence increasing the mass with a factor two doubles the luminosity,
but increasing the distance by a factor two decrease the flux with a
factor four. One should also note that the IC upscattered SD photons
scales slightly softer with mass, hence giving rise to a larger
fraction of gamma-rays in low mass clusters compared to the total
gamma-ray flux. \del{Furthermore, the low mass clusters have a higher flux
from IC upscattered SD photons than high mass clusters which increase
the importance of the SD component in low mass clusters even more
(c.f. Fig.~\ref{fig:IR_comp}).}

@@

{\bf
peak sensitivity of Fermi at these energies due to a combination of effective area background and source spectra...}

In this section we have found that the LP DM models overproduce both
the differential and integrated gamma-ray flux upper limits set by the
Fermi-LAT 18 month data for several clusters. We can use these upper
limits to constrain both the SFE and boost from
substructures. Furthermore, as more sensitive experiments and better
upper limits emerge, we can start ruling out models that give rise to
boost factors that are manifested in the gamma-ray flux. It is
especially interesting is to constrain the SFE, where a boost of 300
is required to explain the excess of electrons and positrons observed
in the vicinity of Earth by PAMELA/Fermi/HESS. In
Fig.~\ref{fig:boost_const} we use the estimated flux from four
clusters to constrain the SFE as well as the minimum mass of
substructures where $M_\rmn{lim}\propto\mvir^{0.226}$. As expected,
the Fornax cluster is the most constraining cluster by far, although
the current upper limits are not good enough to rule out the boost
from substructures or SFE. Assuming that the electron and positron
excess is due to SFE DM, then in order not to overproduce the
differential upper limits set by Fermi-LAT in the energy range
$1-10$~GeV \footnote{We average the LP DM flux in the energy range
  $1-10$~GeV before we compare it to the upper limits.}, we can
constrain the smallest mass of substructures to $M_\rmn{lim}\gtrsim
100\,\msun$. Instead assuming the standard mass of the smallest
substructures of $M_\rmn{lim}\approx 10^{-6}\,\msun$, we can constrain
the boost from SFE to about 2 in Fornax, which corresponds to a
maximum SFE in the MW of about 5. Note that after about six years of
observations with Fermi-LAT, we will be able to rule out the LP model
assuming a substructure boost with a smallest mass of $M_\rmn{lim} =
10^{-6}\,\msun$.

@@

\section{Surface brightness profiles}
\label{sect:spatial}
{\bf should probably be integrated into text; DUSTY STAR FORMING
  GALAXIES IN OUTER PARTS (LITTLE OVERLAP), DUST WOULD NOT SURVIVE. IF
  DUST RAMPRESSURE STRIPPED, SPATIALLIATION PROCESS IN THE HOT ICM FOR
  DYNAMICAL TIME} The large angular extent of clusters on the sky in
combination with the small PSF of most gamma-ray probes ($\sim
0.1$~deg) suggest that we would be able to probe the different spatial
regimes of a cluster. Especially important is the spatial distribution
of the gamma-ray emission as it bias upper limits derived for
spatially extended clusters where a simple profile is often
assumed. In fact when DM substructures are included, the gamma-ray
brightness profile is almost flat and very different from the
gamma-ray distribution following a smooth density profile with a steep
outer brightness slope of $-5$.

>>>>

In supernovae remnants and on scales of galaxies, there are convincing
evidences of non-thermal populations. Especially, in the MW, the
cosmic rays are observed directly as well as indirectly through radio,
X-ray, and gamma-ray emission. On larger scales of the order of few
100~kpc up to Mpc, there are currently a vast number of observations
of radio emission coming from radio mini halos in the center of
cooling flow clusters, radio relics in the periphery of clusters
\cite{2004rcfg.proc..335K} as well as smooth giant radio halos that
have been observed in more than 50 clusters
\cite{2003ASPC..301..143F,2008SSRv..134...93F}. This type of emission
is expected in clusters since the formation process of a galaxy
cluster is a very energetic processes that induces both turbulence as
well as frequently occurring merging and accretion shocks which both
are thought to accelerate relativistic non-thermal protons and radio
emitting electrons to high energies. Thus most of the observed radio
emission in clusters is expected to emerge from cosmic-ray electron
induced radio synchrotron emission. The precise origin of these
electrons, in especially relics and giant radio halos, is still not
settled. One possible scenario for the production of these electrons
are hadronic CR interactions with ambient gas protons which results in
the production of both charged and neutral pions that decay into
electrons, neutrinos, and gamma-rays (see
\cite{1980ApJ...239L..93D,1982AJ.....87.1266V,1999APh....12..169B,2000A&A...362..151D,2001ApJ...559...59M,2003MNRAS.342.1009M,2004A&A...413...17P,2004MNRAS.352...76P,2008MNRAS.385.1211P,2008MNRAS.385.1242P,2009JCAP...09..024K,2010MNRAS.401...47D,2010MNRAS.407.1565D,2010ApJ...722..737K,2010arXiv1011.0729K,2011A&A...527A..99E}
). \footnote{An alternative scenario is the second order turbulent
  re-acceleration through the interaction of a previously injected
  relativistic population of electrons by supernova driven winds or
  active galactic nuclei with plasma waves and
  magneto-turbulence.\cite{1987A&A...182...21S,1993ApJ...406..399G,2004MNRAS.350.1174B,2005MNRAS.363.1173B,2007MNRAS.378..245B,2011MNRAS.410..127B,2009A&A...507..661B}}
Supporting evidence comes from the smoothness of the extended radio
emission that often resembles that observed in thermal
X-rays. Especially since the long cooling time of cosmic ray protons
(CRs) allows for an extended and large population of CRs to build up
over the cluster history \citep{1996SSRv...75..279V,
  1997ApJ...477..560E, 1997ApJ...487..529B}. The production of these
secondaries depend both on the gas and CRp densities in the cluster,
where the CRp density roughly traces the gas outside the core regime
and is slightly enhanced in the center.  This density scaling implies
that clusters are great targets for Cherenkov telescopes with a high
sensitivity for the central parts of nearby clusters. Detecting the
cluster gamma-ray emission is crucial in this respect as it
potentially provides the unique and unambiguous evidence of CRp
populations in clusters through observing the $\pi^0$ bump at about
100~MeV in the spectra.

We adopt the universal spectral and spatial gamma-ray model developed
by Pinzke \& Pfrommer \cite{2010MNRAS.409..449P} to estimate the
emission from decaying $\pi^0$:s that dominates over the IC emission
from primary and secondary electrons above 100~MeV in clusters. The
gamma-ray formalism was derived from high resolution simulations of
clusters of galaxies that included radiative hydrodynamics, star
formation, supernova feedback, and followed CRp physics using a novel
formulation that trace the most important injection and loss processes
self-consistently while accounting for the CRp pressure in the
equation of motion
\cite{2008A&A...481...33J,2007A&A...473...41E,2006MNRAS.367..113P}.
We note that the overall normalization of the CRp and gamma-ray
distribution scales with the maximum acceleration efficiency at
structure formation shock waves. Following recent observations at
supernova remnants \cite{2009Sci...325..719H} as well as theoretical
studies \cite{2005ApJ...620...44K}, we adopt a realistic value of this
parameter and assume that 50\% of the dissipated energy at strong
shocks is injected into CRs while this efficiency rapidly decreases
for weaker shocks \cite{2007A&A...473...41E}. As a caveat we note that
the analytic model is based on simulations that only considered
advective transport of CRs by turbulent gas motions which produces
centrally enhanced profiles. However active CR transport such as CR
diffusion and streaming tends to drive the CR radial profiles towards
being flat, with equal CR number density everywhere. While the CR
streaming is then usually larger than typical advection velocities and
becomes comparable or lower than this only for periods with trans- and
supersonic cluster turbulence during a cluster merger. As a
consequence a bimodality of the CR spatial distribution results with
merging (relaxed) clusters showing a centrally concentrated (flat) CR
energy density profile \cite{2011A&A...527A..99E}.  This translates
into a bimodality of the expected diffuse radio and gamma-ray emission
of clusters, since more centrally concentrated CR will find higher
target densities for hadronic CR proton interactions
\cite{2011A&A...527A..99E}. As a result of this, relaxed clusters
could have a reduced gamma-ray luminosity by up to a factor of five.

@@

We focus in this section on the spectral distribution of gamma-rays
from clusters.  In the leptophilic model, DM annihilation radiation
includes final state radiation and gives rise to substantial amounts of
electrons and positrons that IC upscatter background radiation fields
to high energies. We also consider a few supersymmetric DM models with
a high gamma-ray yield in the form of continuum emission and IC
induced emission. In addition to the annihilating DM, we estimate the
gamma-ray flux induced by shock accelerated CRs.

In Fig.~\ref{fig:diff_BM} we show the differential flux from the
Fornax cluster which is one of the best clusters for indirect DM
searches due to its relative high DM induced fluxes and low CR induced
fluxes. We show the emission for four different supersymmetric BM
models and contrast it to the emission induced by CRs. Comparing this
emission to the differential gamma-ray upper limits set by Fermi-LAT
after 18 months of observations \cite{2010ApJ...717L..71A}, we find
that the upper limits are not violated. The predicted DM flux that is
dominated by the continuum emission from $\Kp$ and $\Ip$ models (shown
in the left panels) are close to the upper limits and can potentially
be probed by the Fermi three-year data. Furthermore, the gamma-ray
signal induced by CRs is expected to be about a factor 5--10 below
the DM continuum flux from the $\Kp$ and $\Ip$ BM models at
10~GeV. However, the IC emission from upscattered CMB and SD photons
is at least a factor 10 lower than the expected flux from CRs above
100~MeV, making it very hard to distinguish such a signal from the
foreground due to the similar spectral index. For energies below
100~MeV we expect IC emission by primary shock accelerated electrons
to be dominating \cite{2010MNRAS.409..449P} over the supersymmetric DM
induced leptons. Hence in clusters, the IC emission from
supersymmetric DM benchmark models be  neglected compared to the
CR pion and DM continuum emission.

@@

One of the main contributions to the expected extended gamma-ray flux
is coming from the IC upscattered photons on various photon background
fields. Depending on the spatial and spectral distribution of
electrons and positrons as well as the photon background field, the
resulting spectral distributions of gamma-ray can differ
greatly. Hence it is interesting to understand which spectral features
and energy regimes  are dominating for the different DM models.

If the enhancement due to substructures in clusters is significant,
then the distribution of electrons and positrons follows the radial
profile of substructures outside the cluster center where the smooth
DM density profile is dominating. In addition, the cooling of the
steady state electron and positron distribution is dominated by SD
photons in the center. Especially since the SD distribution is
centrally peaked and the substructure distribution peaks in the
outskirts around $\rvir$, the overlap between electrons/positrons and
SD photons is small. This results in a suppression of the IC
upscattered SD photons relative to the IC upscattered CMB photons, the
FSR, and the continuum emission. However, if substructures are only
marginally dominating over the smooth distribution in a cluster, the
SD component becomes relatively more important.

In Fig.~\ref{fig:IR_comp} we show the total IC emission as well as its
individual contributions from different IC upscattered radiation
fields. The left panel shows the gamma-ray emission from the LP model
and the right panel from the $\Kp$ BM model, and for comparison we
overplot the emission expected from the CRs. Due to the flat electron
and positron spectrum resulting from the LP DM model and the smaller
mean energy of the CMB compared to the SD, we find that the
upscattered CMB photons dominate the total DM IC emission in the
energy regime below 100~GeV, while the SD dominate above this
energy. For the BM models this transition energy is shifted towards
smaller energies since the electron and positron spectrum has a
steeper spectral distribution (see Fig.~\ref{fig:q_DM}). At the
highest gamma-ray energies of about 100~GeV and above, the IC from
starlight steepens because it probes the high energy tail of the
electrons and more importantly, it suffers from the Klein-Nishina
suppression. When substructures are present, most of their flux
resides in the outer parts of clusters. However, if we remove the
boost from substructures, the density profile of electrons and
positrons is more centrally peaked, and the relative importance of the
IC upscattered SD increases by a factor $\sim 30$.  A larger fraction
of electrons (in the core) will now also cool by Compton-upscattering
SD photons which suppresses the IC-upscattered CMB light.  For
comparison, we include the contribution of IC upscattered SD photons
and to bracket the uncertainty in the SD model, with a smaller amount of
dust. In this low-dust model, we have reduced the energy in dust by a
factor 10. However, we find that the IC SD flux only decreases by a
factor 3 since the IC cooling of the electrons and positrons also
decreases. Note that for a smaller boost from substructures, the
effect from cooling is smaller.

 \del{This
  suppression is also seen in the relative normalization of the flux
  where the IC from both CMB and dust photons are lower compared to
  the starlight simply because the starlight has a higher energy,
  hence it is upscattered by more abundant low energy electrons and
  positrons (c.f. Fig.~\ref{fig:q_DM}).}
\del{Even though the SD component seems to be suppressed compared to
  the CMB at energies around 100~MeV, it has a relative large impact
  on the steady state electron spectrum. If we remove the SD cooling,
  we find a factor two larger IC flux at low energies for a cluster of
  the size of Fornax. For more massive clusters we expect a larger IC
  flux from the SD component, but a smaller contribution from the CMB
  because of the more effective cooling.}

Because of the large total boost factor for the LP model ($\sim
10^4$), we overproduce the upper gamma-ray limit in the $1-10$~GeV
energy interval set by Fermi-LAT by about a factor 100. This strongly
constrains the boost from both substructures and the Sommerfeld
enhancement. Additionally, these constraints might improve with future
more sensitive Cherenkov telescopes such as CTA. However, considering
that an indirect detection of DM in clusters relies on the boost from
substructures whose main contribution comes from the periphery, we
conclude that the wide angular extent of clusters on the sky in
gamma-rays suggest that these sources are not ideal for IACTs since
their sensitivity drops linearly with source extension.

@@

It is also interesting to compare the total contribution from the LP
model, the brightest BM model ($\Kp$), and the CRp induced
emission. In Fig.~\ref{fig:flux_int} we show the integrated flux from
Fornax for our different gamma-ray models and compare it to the
unresolved integrated flux upper limit on Fornax set by Fermi-LAT
where they averaged the flux over the energy range $0.2-100\,\gev$
assuming a spectral index of 2. Again, the annihilation flux in the LP
model is in conflict with the upper limits by Fermi-LAT, although only
by a factor of 10 which is about a order magnitude less constraining
than the differential flux in the energy range $1-10$~GeV. As shown in
this figure, the LP model is dominating the entire gamma-ray energy
range up to the DM rest mass energy in this model of about
1~TeV. Finally, we note that the flux the DM BM $\Kp$ model is above
the predicted emission from the CRs in the $1-100$~GeV energy
regime. Hence, the prospects for detecting the $\Kp$ BM model over the
expected gamma-ray background induced by CRs looks promising.

@@

We continue by comparing the estimated differential flux from Fornax
to three other clusters in Fig.~\ref{fig:clu_comp}; the close by and
well studied Virgo cluster, the X-ray bright Perseus clusters, and the
massive merging Coma cluster. We find high DM induced gamma-ray fluxes
from both the Fornax and Virgo clusters, which highlight them as
promising targets for indirect DM searches. Note, however, that we
have used spatially unresolved upper limits, which is quite optimistic
for the Virgo cluster which has an angular extended of 7 degrees. It
is quite striking how constraining the upper limits for the Fornax
cluster are compared to the other clusters, especially in the
$1-10$~GeV energy regime. At these energies, Fermi-LAT has its peak
sensitivity due to a combination of increasing effective area and
decreasing source spectra as a function of energy. The upper limits
for Virgo and Perseus are unfortunately background contaminated by AGN
activity from M87 and NGC 1267, respectively, and do not gain much
from the increased sensitivity. Note that the high CR-induced flux in
Coma and Perseus is of the same magnitude as the DM flux from LP
models, making it difficult for indirect DM searches in those
clusters. Conversely is the Fornax cluster a great target for indirect
DM studies because the relative low gamma-ray flux from CRs, absents
of an active AGN, and high DM gamma-ray flux. The Fermi-LAT three-year
data will be able to constrain the smallest mass of substructures
using the DM BM models.

What is the figure of merit for selecting the most promising cluster
targets for indirect DM searches? To this end, we employ the
luminosity-to-mass scaling relations. The gamma-ray luminosity from
the smooth density profile is given by \cite{2009PhRvL.103r1302P}
\begin{equation}
L_{\gamma,\sm} \propto \int \dd V \rho(r)^2 \propto \frac{M_{200}\,c^3}
{\left[\log\left(1+c\right)-c/(1+c)\right]^2} \propto \mvir^{0.83}\,,
\end{equation}
and the total DM luminosity that includes boost factors for the LP and
the BM models are given by
\begin{eqnarray}
\label{eq:LP_scaling}
L_{\gamma,\rmn{LP}} &=& L_{\gamma,\rmn{smLP}} \B_\rmn{LP} \propto \frac{\mvir^{0.73}}{D_\rmn{lum}^2},\\
\label{eq:BM_scaling}
L_{\gamma,\rmn{BM}} &=& L_{\gamma,\rmn{smBM}} \B_\rmn{BM} \propto  \frac{\mvir^{1.06}}{D_\rmn{lum}^2},\\
\B_\rmn{LP}&=&\B_\sfe \B_\sub \propto \mvir^{-1/3}\mvir^{0.23} \propto
 \mvir^{-0.1}\,,\\
\B_\rmn{BM}&=&\B_\sub \propto \mvir^{0.23}\,.
\end{eqnarray}
Hence increasing the cluster mass in the BM model by a factor two more
than doubles the luminosity, but increasing the luminosity distance,
$D_\rmn{lum}$, by a factor two decrease the flux by a factor four. One
should also note that the IC upscattered SD photons scales slightly
softer with mass, hence giving rise to a larger fraction of gamma-rays
in low mass clusters compared to the total gamma-ray flux.

\del{Furthermore, the low mass clusters have a higher flux from IC
  upscattered SD photons than high mass clusters which increase the
  importance of the SD component in low mass clusters even more
  (c.f. Fig.~\ref{fig:IR_comp}).}

@@

In this section we have found that the LP DM models overproduces both
the differential and integrated gamma-ray flux upper limits set by the
Fermi-LAT 18 month data for several clusters. We can use these upper
limits to constrain the boosts due to SFE and
substructures. Furthermore, as more sensitive experiments and better
upper limits emerge, we can start ruling out models that give rise to
boost factors that are manifested in the gamma-ray flux. It is
especially interesting is to constrain the SFE, where a boost of the
particle physics cross section of 300 is required to explain the
excess of electrons and positrons observed in the vicinity of Earth by
PAMELA/Fermi/HESS. In Fig.~\ref{fig:boost_const} we use the estimated
flux of four clusters to constrain the SFE as well as the minimum mass
of substructures where $M_\rmn{lim}\propto\mvir^{0.226}$. As expected,
the Fornax cluster is by far the most constraining cluster, although
the current upper limits are not good enough to rule out the boost
from substructures or SFE. Assuming that the PAMELA positron excess is
due to SFE DM, then in order not to overproduce the differential upper
limits set by Fermi-LAT in the energy range $1-10$~GeV \footnote{We
  average the LP DM flux in the energy range $1-10$~GeV before we
  compare it to the upper limits.}, we can constrain the smallest mass
of substructures to $M_\rmn{lim}\gtrsim 100\,\msun$. Instead assuming
the standard mass of the smallest substructures of $M_\rmn{lim}\approx
10^{-6}\,\msun$, we can constrain the boost from SFE to about 2 in
Fornax, which corresponds to a maximum SFE in the MW of about 5. Note
that after about six years of observations with Fermi-LAT, we will be
able to rule out the LP model assuming a substructure boost with a
smallest mass of $M_\rmn{lim} = 10^{-6}\,\msun$.

@@

\section{Surface brightness profiles}
\label{sect:spatial}

\del{\bf should probably be integrated into text; DUSTY STAR FORMING
  GALAXIES IN OUTER PARTS (LITTLE OVERLAP), DUST WOULD NOT SURVIVE. IF
  DUST RAMPRESSURE STRIPPED, SPATIALLIATION PROCESS IN THE HOT ICM FOR
  DYNAMICAL TIME}

The large angular extent of clusters on the sky in combination with
the small PSF of most gamma-ray probes ($\sim 0.1$~deg) suggest that
we would be able to probe the different spatial regimes of a
cluster. Especially important is the spatial distribution of the
gamma-ray emission as it bias upper limits derived for spatially
extended clusters where a simple profile is often assumed. In fact, we
will show that if DM substructures are present, the gamma-ray
brightness profile is almost flat and very different from the
gamma-ray distribution following a smooth density profile with a steep
outer surface brightness slope of $-5$.
